---
title: "R/RStudio (Part 1). Data wrangling & visualization (CSB chap 9)"
author: "Eamen Ho & Weigang Qiu"
date: "2023-03-08"
output:
#  word_document: default
#  pdf_document: default
  html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```


## 9.2 Welcome to the tidyverse

```{r}
#install.packages("tidyverse") # install tidyverse package
library(tidyverse) # load package
#tidyverse_update() # update all packages 
tidyverse_packages() # list all packages belonging to the package
```

### 9.2.1 Reading Data

```{r}
#setwd("data_wrangling/sandbox/") # set to your working directory
# Data from: Fauchald P, Park T, Tømmervik H, Myneni R, Hausner VH (2017) Arctic greening from warming promotes declines in caribou populations. Science Advances 3(4): e1601365. http://dx.doi.org/10.1126/sciadv.16013
popsize <- read_tsv("data_wrangling/data/FauchaldEtAl2017/pop_size.csv") # read the tsv file 
```

### 9.2.2 Tibbles

```{r}
popsize # shows the data structure of the file popsize (tibble)
head(popsize, 3) # shows the first 3 lines
tail(popsize, 3) # shows the last 3 lines
#View(popsize) # open the data in a spreadsheet-like environment
glimpse(popsize) # brief summary of the structure of the data
dim(popsize) # dimensions of the data (# of rows, # of columns)
nrow(popsize) # number of rows
ncol(popsize) # number of columns
```

## 9.3 Selecting and Manipulating Data

```{r}
# read vegatation data:
# NDVI: Landsat Normalized Difference Vegetation Index (NDVI) is used to quantify vegetation greenness
ndvi <- read_tsv("data_wrangling/data/FauchaldEtAl2017/ndvi.csv")
head(ndvi)
```

### 9.3.1 Subsetting Data

```{r}
# extract only specific columns by name (e.g., Herd and NDVI_May)
select(ndvi, Herd, NDVI_May) 
# use a colon to include all columns in between
select(ndvi, Herd:NDVI_May)
# use a minus sign to exclude specific columns
select(ndvi, -Herd, -Year)
# you can combine the two features
select(ndvi, -(Year:NDVI_May))
# use a regular expression for the name of the column > select(ndvi, matches("NDV"))

# select rows with value "WAH" for variable Herd
filter(popsize, Herd == "WAH")
# select rows for years 1970 to 1980
filter(popsize, Year >= 1970, Year <= 1980)
# select specific, nonconsecutive years
filter(popsize, Year %in% c(1970, 1980, 1990))

# select rows according to row number
# e.g., everything between rows 20 and 30
slice(popsize, 20:30)
# top 10 rows when ordered by Pop_Size
top_n(popsize, 10, Pop_Size)
# take the 15 rows with smallest Pop_Size 
top_n(popsize, 15, desc(Pop_Size))
# extract 5 rows at random
sample_n(popsize, 5)
# 2% of the rows at random
sample_frac(popsize, 0.02)
```

### 9.3.2 Pipelines

```{r}
# an example of nested base R code
unique(popsize[order(popsize$Herd), 1])

# first, take only column Herd
select(popsize, Herd)
# second, remove repeated values, using distinct 
distinct(select(popsize, Herd))
# finally, sort the data using arrange
arrange(distinct(select(popsize, Herd)), Herd)
```

The tidyverse package provides a "pipeline" operator (%), similar to the vertical pipe (\|) in unix.

```{r}
popsize %>% 
  select(Herd) %>% 
  distinct() %>% 
  arrange(Herd)

# save into a new object
herds <- popsize %>% 
  select(Herd) %>% 
  distinct() %>% 
  arrange(Herd)
```

## Intermezzo 9.1

(a) Extract the unique Years in popsize and order them.
(b) Find the row in ndvi with the largest NDVI_May.
(c) List the three years with the largest Pop_Size for the Herd called WAH.Perform this operation in two different ways.

## 9.3.3 Renaming Columns

```{r}
popsize %>% 
  rename(h = Herd) # rename a column
```

### 9.3.4 Adding Variables

To add a new variable whose content is a function of some other columns, use mutate.

For example, suppose you want to add a new column to the ndvi tibble, called meanNDVI, contaiing the average of the values in May and those in June/August:

```{r}
ndvi %>% 
  mutate(meanNDVI = (NDVI_May + NDVI_June_August)/ 2) %>% 
  head(4)
```

If you use transmute instead of mutate, you retain only the column9s) containing the results:

```{r}
ndvi %>% transmute(meanNDVI = (NDVI_May + NDVI_June_August) / 2) %>% head(4)
```

## Counting and Computing Statistics

### 9.4.1 Summarize Data

You can use the operator summarise to create summaries of the data

```{r}
ndvi %>% 
  summarise(mean_May = mean(NDVI_May)) # average NDVI_May for the whole data

ndvi %>% 
  summarise(mean_May = mean(NDVI_May), sd_May = sd(NDVI_May),
                   median_May = median(NDVI_May)) # can compute many statistics at once
```

### 9.4.2 Grouping Data

You can perform operations like mutate and summarise on grouped data using the dplyr package

```{r}
popsize %>% 
  group_by(Herd) %>% 
  summarise(avgPS = mean(Pop_Size), minPS = min(Pop_Size), maxPS = max(Pop_Size)) %>% 
  arrange(Herd) # compute the average, min, max population sizes for each herd, taken across all years

# use tally to count the number of rows belonging to each group
popsize %>% 
  group_by(Herd) %>% 
  tally() %>% 
  arrange(Herd)

# the function n counts rows
popsize %>% 
  group_by(Herd) %>%
  summarise(tot = n()) %>% 
  arrange(Herd)
```

To show how you can use mutate (or transmute) on grouped data, we are going to compute a z-score for the population size of each herd in each year.

```{r}
popsize %>% 
  group_by(Herd) %>% 
  mutate(zscore = (Pop_Size - mean(Pop_Size)) / sd(Pop_Size))
```

## Intermezzo 9.2

(a) Compute the average Pop_Size for each Herd in popsize.
(b) Identify the Herd with the largest standard deviation for NDVI_May.
(c) Add a new column containing the population size relative to the mean population size (i.e., Relative_Pop should be 1 when the population in that year is exactly as large as the mean population, a value of 2 if the population is twice the mean, 0.5 if it's half of the mean, etc.).

## 9.6 Data Visualization

### 9.6.2 The Structure of a Plot

```{r}
# load the library
library(tidyverse)
# read the data
popsize <- read_tsv("data_wrangling/data/FauchaldEtAl2017/pop_size.csv")
ndvi <- read_tsv("data_wrangling/data/FauchaldEtAl2017/ndvi.csv") 
seaice <- read_tsv("data_wrangling/data/FauchaldEtAl2017/sea_ice.csv") 
snow <- read_tsv("data_wrangling/data/FauchaldEtAl2017/snow.csv")
# convert to tidy form
seaice <- seaice %>% gather(Month, Cover, 3:14)
```

Let's build our first graph by adding the elements described in the the grammar of graphics, one by one.

```{r}
ggplot(data = popsize) #produces an empty graph
```

We can associate the x-axis with the Year, the y-axis with Pop_Size, and the color of the points with Herd by typing:

```{r}
ggplot(data = popsize) + 
  aes(x = Year, y = Pop_Size, colour = Herd)
```

Notice that we have added the aes using the + sign, which has been repur- posed to collate the various plotting commands. Still we don't have a full graph, because we need to specify a "geometry." For example, let's draw points connected by segments:

```{r}
ggplot(data = popsize) + 
  aes(x = Year, y = Pop_Size, colour = Herd) + 
  geom_point() + geom_line()
```

### 9.6.3 Plotting Frequency Distribution of One Continuous Variable

```{r}
#plot a histogram 
ggplot(data = ndvi) + 
  aes(x = NDVI_May) + 
  geom_histogram() #each bar represents the number pf data points (count) that fall into a certain data range (bin)

# nested or "unrolled"
# these three commands produce the same graph: 
ggplot(data = ndvi) + 
  aes(x = NDVI_May) + 
  geom_histogram()

ggplot(data = ndvi, aes(x = NDVI_May)) + 
  geom_histogram() 

ggplot(data = ndvi) + 
  geom_histogram(aes(x = NDVI_May))

#interpolate the histogram and produce a density plot
ggplot(data = ndvi) + aes(x = NDVI_May) + geom_density()
```

### 9.6.4 Box Plots and Violin Plots

Box plots and violin plots are excellent for providing information on the distribution of your data. A box plot marks the median (50th percentile) of the values as a line in the middle of a box enclosing half of your data (from the 25th to the 75th percentile). The lines extending from the box (whiskers) rep- resent 1.5 × the length of the box (or interquartile range, IQR). Data points falling beyond the whiskers are drawn separately as outliers.

Violin plots add information on the distribution of data (i.e., where a violin plot is wider, more data points cluster---you can think of a violin plot as the union of two mirrored density plots).

```{r}
pl <- ggplot(data = ndvi) + aes(x = Herd, y = NDVI_May) 
pl + geom_boxplot()
pl + geom_violin()

pl + geom_boxplot() + aes(fill = Herd) # change the color for the boxes, set the aesthetic fill

```

### 9.6.5 Bar Plots

```{r}
ggplot(data = seaice %>% filter(Herd == "WAH")) + 
  aes(x = Year) + 
  geom_bar() # shows that we have 12 records per year for all the years between 1979 and 2014

# When we want the height of the bars to represent a value in a column (instead of a count), we can use geom_col
ggplot(data = seaice %>% 
         filter(Herd == "WAH", Year == 1990)) + 
  aes(x = Month, y = Cover) + geom_col()

seaice$Month <- factor(seaice$Month, month.abb) 
ggplot(data = seaice %>% 
         filter(Herd == "WAH", Year == 1990)) + 
  aes(x = Month, y = Cover) + geom_col()
```

### 9.6.6 Scatter Plots

Scatter plots show the relationship between two continuous variables. Let's plot the population dynamics of the herd WAH in time:

```{r}
pl <- ggplot(data = popsize %>% filter(Herd == "WAH")) + 
  aes(x = Year, y = Pop_Size) + 
  geom_point()

show(pl) # The function show is used to display the plot stored in pl

pl + geom_smooth()

# add a regression line
pl + geom_smooth(method = "lm") # use a polynomial regression
pl + geom_smooth(method = "lm", formula = y ~ poly(x, 3), se = FALSE) # By setting se = FALSE, we suppress plotting the standard error
```

### 9.6.7 Plotting Experimental Errors

The standard deviation (SD) represents the dispersion of the data,4 the standard error of the mean (SEM) and confidence intervals (CI) report the certainty of the estimate of a value (e.g.,certainty in estimating the mean).

```{r}
#calculate summary stats and errors for herds GRH and PCH
stats <- popsize %>% 
  filter(Herd %in% c("GRH", "PCH")) %>% 
  group_by(Herd) %>% 
  summarise(meanPopSize= mean(Pop_Size), SD = sd(Pop_Size), N = n(), SEM = SD/sqrt(N), CI = SEM * qt(0.975, N-1))

# plot mean population size as bar plot
ggplot(data = stats) + 
  aes(x = Herd, y = meanPopSize) + 
  geom_col()

# set up aesthetic mapping for confidence intervals
limits <- aes(ymax = stats$meanPopSize + stats$CI, ymin = stats$meanPopSize - stats$CI)
# plot including confidence intervals
ggplot(data = stats) + 
  aes(x = Herd, y = meanPopSize) + geom_col() + 
  geom_errorbar(limits, width = .3)
```

### Intermezzo 9.5

(a) Explore the data set snow and plot the Perc_snowcover versus Week_snowmelt. Choose an appropriate geom.
(b) For each year, compute the average Perc_snowcover and Week_snowmelt and create a scatter plot.
(c) Produce a box plot of the Pop_Size for each Herd.
(d) Create a scatter plot showing the average population size (taken across herds) including standard deviation, for the years 2008 through 2014.

### 9.6.8 Scales

Scales are used to modify the x- and y-axes, as well as to determine how colors, sizes, and shapes behave. You can set a scale for each aesthetic mapping. Here are a few other aesthetic mappings that you can modify using the corresponding scales:

insert pic pg 326

```{r}
pl <- ggplot(data = popsize, aes(x = Herd, y = Pop_Size, fill = Herd)) +
  geom_boxplot()

show(pl)

# choose a palette from ColorBrewer
pl + scale_fill_brewer(palette = "Set3")
# palette based on hue
pl + scale_fill_hue()
# manually set values and rename the legend
pl + scale_fill_manual(values = rainbow(11), name = "aaa")

#Similarly, you can set other scales such as size or colour. Take the graph
pl <- ggplot(data = seaice %>% filter(Herd == "BEV")) + 
  aes(x = Year, y = Month, colour = Cover, size = Cover) + 
  geom_point()
show(pl)

pl + 
  scale_colour_gradient(high = "white", low = "red")
```

### 9.6.9 Faceting

Sometimes you want to split data according to one or more variables, but still plot the resulting panels in one plot for easy comparison. In ggplot2 this operation is called faceting. Two functions are available to facet data: facet_wrap and facet_grid. While facet_wrap arranges panels sequentially and omits empty panels when a certain combination of variables is not available, facet_grid arranges all variable combinations in a grid regardless of data availability.

```{r}
seaice %>% 
  filter(Herd %in% c("WAH", "BAT")) %>% 
  filter(Year %in% c(1980, 1990, 2000, 2010)) %>% 
  ggplot(aes(x = Month, y = Cover)) + 
  geom_col() + 
  facet_grid(Year~Herd)
```

In a facet_grid, the panels have common x- and y-axes, and are ordered according to up to two variables (in our example the variable Year is mapped into the rows of the grid, and Herd into the columns). Notice that sea-ice cover in September decreased dramatically over four decades. You can use facet_wrap when you want simply to fit several panels on the page, but the ordering of the panels has no particular meaning. For example, the sea-ice cover in 2010 for all populations can be plotted using

```{r}
ggplot(data = seaice %>% filter(Year == 2010)) + 
  aes(x = Month, y = Cover) + 
  geom_col() + 
  facet_wrap(~Herd)

# One nice feature of facet_wrap is that it allows for the ranges of the x- and y-axes to change between the panels (very useful when they vary considerably): 
ggplot(data = seaice %>% filter(Year == 2010)) + aes(x = Month, y = Cover) + geom_col() + facet_wrap(~Herd, scales = "free")
```

### 9.6.10 Labels

We can change the labels of the graphs via xlab, ylab, and ggtitle:

```{r}
pl <- ggplot(data = popsize) + 
  aes(x = Year, y = Pop_Size) + 
  geom_point()
pl + xlab("Year")
pl + ylab("Population Size")
pl + ggtitle("Population Dynamics")
```

### 9.6.11 Legends

By default, ggplot2 will place all the legends on the right of your plot. You can move them all to another position by specifying the legend.position option:

```{r}
pl <- ggplot(data = popsize) + aes(x = Herd, y = Pop_Size, fill = Herd) +
  geom_boxplot()
# default
show(pl)
# move legend
pl + theme(legend.position = "bottom")
pl + theme(legend.position = "top")
# remove legend
pl + theme(legend.position = "none")
```

Each legend comes with a guide. You can use the command guide to change the title of a legend, alter its features, or to suppress it altogether. For example,

```{r}
pl <- ggplot(data = popsize) + 
  aes(x = Year, y = Pop_Size, colour = Herd, alpha = sqrt(Pop_Size)) +
  geom_point()

show(pl)

pl + guides(colour = guide_legend(nrow = 4,title = "herd"), alpha = guide_legend(direction = "horizontal", title = "al"))

# suppress only one legend
pl + guides(colour = "none")
```

### 9.6.12 Themes

Once we have built a plot we like, we can change its general appearance by choosing a different theme. Themes allow you to quickly and consistently adapt your plots to printing requirements, etc. Here are some examples:

```{r}
# load data if not done for intermezzo
snow <- read_tsv("data_wrangling/data/FauchaldEtAl2017/snow.csv") 
pl <- ggplot(data = snow %>% filter(Herd == "CAH"), aes(y = Week_snowmelt, x = Perc_snowcover)) + 
  geom_point()
# default
show(pl)
# black and white (light background) pl + theme_bw()
# line draw
pl + theme_linedraw()
# minimalist theme
pl + theme_minimal()
```

The package ggthemes provides several extra themes, including themes that mimic the style of popular publications:

```{r}
#install.packages("ggthemes")
library(ggthemes)
# Wall Street Journal
show(pl + theme_wsj())
# five thirty-eight
show(pl + theme_fivethirtyeight())
```

### 9.6.12 Setting a Feature

Sometimes we want to set one of the features of the graph, rather than mapping some data to it. For example, we want to have a certain size, color, shape, or transparency for all the points. To do so, we need to set the value outside the aesthetics:

```{r}
# use color as an aesthetic mapping, associated with Herd
pl <- ggplot(data = popsize) + aes(x = Year, y = Pop_Size, colour = Herd) + geom_point()
# set color to be red for all points
pl <- ggplot(data = popsize) + aes(x = Year, y = Pop_Size) + geom_point(colour = "red")
```

### 9.6.12 Saving

Use the function ggsave to save a plot:

```{r}
ggsave(filename = "data_wrangling/sandbox/test.pdf", plot = pl, width = 3, height = 4)
```

### Intermezzo 9.6

(a) For each herd (facet), produce a graph showing the histogram of the week in which ground snow melts.
(b) A geometry we didn't explore is geom_tile, which can be used to pro- duce heat maps. Explore its documentationa and draw a heat map having Year as the x-axis, Herd as the y-axis, and Week_snowmelt as the color of each cell.
(c) Produce a scatter plot showing the Perc_snowcover vs. Week\_ snowmelt. Facet using Herd, and add a smoothing line to each panel.

## 9.7 Tips & Tricks

If a column name is a number, or contains white spaces, you can refer to it in the code by using back ticks as quotes:

```{r}
popsize %>% 
  filter(Year > 1979, Year < 1985) %>% 
  spread(Year, Pop_Size) %>% select(Herd, `1980`)
```

Sometimes, you need to remove the grouping information before proceeding with your pipeline. To remove all the information on groups from a tibble, pass it to ungroup.

```{r}
# to see the difference, run the code with # and without ungroup
popsize %>% 
  group_by(Herd, Year) %>% 
  tally() %>% 
  ungroup() %>% 
  summarise(n = sum(n))
```

Operations by row. We have performed operations involving columns (or subsets of columns). When you want to perform an operation by row you can use rowwise. Suppose you want to generate a new column that contains the highest values of NDVI_May and NDVI_June_August, respectively, for each year and herd. The following does not work:

```{r}
ndvi %>% 
  mutate(maxndvi = max(NDVI_May,NDVI_June_August)) %>% 
  head(4)
```

This is because the maximum is taken over all the values in each column. In such cases, you can apply the function row by row:

```{r}
ndvi %>% 
  rowwise() %>% mutate(maxndvi = max(NDVI_May, NDVI_June_August)) %>% 
  head(4)
```
